---
title: "Problem Set 3. Due Thurs March 2 5pm"
author: "Eric Prince"
date: "Feb 23, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "#>")
```

# Submission

1. Fork this repository to your own account
1. Make changes to the `README.Rmd` file (including the author field above).
1. Knit the file to HTML format and add it to your git repository (`git add`)
1. Submit a pull request with your Rmd and knitted HTML files.

# Overview 

You will examine a [time-course of mRNA expression][1] of barley powdery mildew fungus in immunocompromised plants (*Arabidopsis*). The fungus Blumeria graminis f. sp. hordei (called *Bgh* from now on)

Counts were generated using the [`maSigPro` software][3] and deposited in [NCBI GEO][2].

[1]: http://www.ncbi.nlm.nih.gov/pubmed/23175786
[2]: http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE43163  
[3]: http://bioinformatics.oxfordjournals.org/content/30/18/2598.full

Some important identifiers for the data set:

| Name | Description |
| ---- | ----------- |
| `A6` | *Bgh* isolate expressing other AVRA effectors |
| `K1` | *Bgh* isolate expressing the cognate AVRA1 effector for MLA1|
| `pps` | *Arabidopsis* plants |
| `B12` | *Arabidopsis* plants expressing MLA1-HA|

We will download tables from GEO containing counts of mRNA abundance from both the fungal samples.

# Raw data

First we need load the data.

```{r load_raw_data, message = FALSE}

# Load necessary packages
    pkgs <- c("tidyverse",
              "magrittr",
              "stringr",
              "reshape2",
              "cowplot")
    pacman::p_load(pkgs, character.only = TRUE)

# bgh data
bgh_url <- 'http://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE43163&format=file&file=GSE43163_CompleteCountTable_Bgh.txt.gz'

# 2 info and blank lines at top, skip them
raw_data <- read_tsv(bgh_url, skip = 2)

# the header for the first column is "NA", set it manually
names(raw_data)[1] <- 'gene.name'

raw_data <- as_data_frame(raw_data)
```

# Problems


## Problem 1

Now that the raw data are loaded, your first task is to tidy the data with `tidyr`. The `cleaned_data` should look something like this:

```
# A tibble: 310,896 × 6
             gene.name plant fungus time.point   rep value
*                <chr> <chr>  <chr>      <chr> <chr> <int>
1             bgh04079   B12     A6       6hpi     1    13
2             bgh01634   B12     A6       6hpi     1    31
3  bghG000012000001001   B12     A6       6hpi     1   121
4  bghG000012000002001   B12     A6       6hpi     1     3
5             bgh00757   B12     A6       6hpi     1   253
6             bgh01273   B12     A6       6hpi     1    45
7             bgh01274   B12     A6       6hpi     1    32
8             bgh01277   B12     A6       6hpi     1     9
9             bgh06140   B12     A6       6hpi     1    47
10            bgh05774   B12     A6       6hpi     1    42
# ... with 310,886 more rows
```

The key steps are to `gather()` the data into key-value pairs and then `separate()` the information from the key column into new columns.

```{r problem_1}

raw_data %>%
  gather(key = tmp, value = value, -gene.name) %>%
  separate(col = tmp, into = c("plant", "fungus", "time.point", "rep"), sep = "_") %>%
  as_tibble() -> cleaned_data
cleaned_data

# alternatively,
# 
# raw_data %>%
#   melt(id.vars = "gene.name") %>%
#   separate(variable, into = c("plant", "fungus", "time.point", "rep"), sep = "_") %>%
# as_tibble() -> cleaned_data
# cleaned_data

# which happens to be a little quicker
#
# > system.time(raw_data %>%
# +                 gather(key = tmp, value = value, -gene.name) %>%
# +                 separate(col = tmp, into = c("plant", "fungus", "time.point", "rep"), sep = "_") %>%
# +                 as_tibble() -> cleaned_data)
#    user  system elapsed 
#   0.571   0.018   0.603 
#
# > system.time(raw_data %>%
# +                 melt(id.vars = "gene.name") %>%
# +                 separate(variable, into = c("plant", "fungus", "time.point", "rep"), sep = "_") %>%
# +                 as_tibble() -> cleaned_data)
#    user  system elapsed 
#   0.377   0.011   0.390 

```

## Problem 2

You need to translate what you learned on the command line into the R world.

Which plant has the highest expression of any gene in the 6hpi time point?

```{bash, eval=FALSE}
raw_data | awk '$4 == "6hpi"' | sort -k6nr | head -n 1 | cut -f2
```

```{r}
cleaned_data %>%
  filter(time.point == "6hpi") %>%
  arrange(desc(value)) %>%
  select(plant) %>%
  slice(1)
```

Which plant / fungus pair has the highest expression in the 18hpi time point?

```{bash, eval=FALSE}
raw_data | awk '$4 == "18hpi"' | sort -k6nr | head -n 1 | cut -f2,3
```

```{r}
cleaned_data %>%
  filter(time.point == "18hpi") %>%
  arrange(desc(value)) %>%
  select(plant, fungus) %>%
  slice(1)
```

## Problem 3

Identify the cltop 3 most consistently differentially expressed genes between the earliest and latest time points for each combination of `plant` and `fungus` strains.

- "Differential expression"" is the difference between `value` (i.e., gene expression level) between time points.

- "Consistency" is the smallest variance in `value` between replicates.

### Strategy

1. Create a new table from the cleaned data by moving each `hpi` value to a new column name, with counts for each in the column (hint: use a `tidyr` verb). It is helpful to reformat the `hpi` values by converting from e.g. `6hpi` to `hpi.6`. You can use `mutate` to do this, i.e.:

```{r reformat_time, eval=FALSE}
cleaned_data %>%
  mutate(time.value = str_replace(time.point, 'hpi', ''),
         time.hpi = str_c('hpi.', time.value))
```

2. Create a new column containing the expression difference between the relevant time points.

3. Calculate summary statistics (mean and variance) of the expression differences by grouping (hint) the `gene.name`, `plant`, and `virus` columns.

4. Sort by these statistics and use the `dplyr` verb `slice` to pull the ones you want (i.e., the top 3). Note you will have to remove `gene.name` from the grouping so that sorting works.

```{r problem_3}

cleaned_data %>%
  spread(time.point, value) %>%
  mutate(diff = `24hpi` - `6hpi`) %>%
  group_by(gene.name, plant, fungus) -> cleanr_data
  
  cleanr_data %>%
    summarise(mean(diff),
              var(diff)) %>%
    filter(`mean(diff)` != 0 & `var(diff)` != 0) %>%
    arrange(`var(diff)`,
            desc(`mean(diff)`)) %>%
    group_by(plant, fungus) %>%
    slice(1:3) -> top.hits
top.hits

```

## Problem 4

Now examine the above final data frame above and write a few sentences putting inline code in least 3 places. For example, There are `r nrow(mtcars)` rows of data in `mtcars`.

Starting from a population `r nrow(cleaned_data)` unique mRNA observations, we identified the top three most consistently differentially expressed genes over the time-course of the experiment.  The top hit had a mean differential expression value of `r max(top.hits[,4])` and it belonged to `r top.hits[max(top.hits[,4]),1]` which was found in the `r top.hits[max(top.hits[,4]),2]` plant and `r top.hits[max(top.hits[,4]),3]` fungus combination.

## Problem 5

Plot the expression (`value`) by time (`hpi`) for the above genes. Format the plot as follows:

1. Plot each point.
2. Connect the points with a smooth line 
3. Create a facet for each `plant` / `bgh` pair 
4. Assign a different color to each gene 
5. Assign a different shape to each replicate 
6. Add a descriptive title to the plot 
7. Use the "black & white" theme

Extra credit: add error bars to the plot (use `geom_errorbar`).

```{r problem_5}

cleaned_data %>%
    mutate(time.value = str_replace(time.point, "hpi", ""),
           time.hpi = str_c("hpi.", time.value)) %>%
    select(1:3,5:7) %>%
    merge(., top.hits) %>%
    arrange(as.numeric(time.value)) -> p.tbl
  
  p <- ggplot(p.tbl, 
              aes(x=factor(time.value, levels = unique(time.value)),
                  y = value, 
                  shape = rep)) +
    geom_point() + 
    stat_summary(aes(y= value,
                     group = gene.name,
                     color = gene.name),
                     fun.y = mean,
                     geom = "line") +
    theme_cowplot() +
    facet_wrap(plant ~ fungus) +
    labs(title = "Timecourse mRNA Expression of \n Barley Powdery Mildew Fungus in Arabidopsis",
         x = "HPI",
         y = "Expression Value",
         color =  "Gene Name")
  p

```
